Appendix

### 2025-10-29 manuscript_rmarkdown/manuscript_appendix.tex
Updates to text.

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  a4paper,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{placeins}
\newtheorem{definition}{Definition}
\DeclareMathOperator*{\argmax}{argmax}
\DeclareMathOperator*{\argmin}{argmin}
\DeclareMathOperator{\sgn}{sgn}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Appendix},
  pdfauthor={Alex Zwanenburg, Steffen Löck},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Appendix}
\author{Alex Zwanenburg, Steffen Löck}
\date{2025-04-11}

@@

The parameters of these power transformations can be optimised based by
maximising the log-likelihood function, under the assumption that the
transformed feature \(\phi^{\lambda, x_0, s} (\mathbf{X})\) follows a
normal distribution. The log-likelihood functions for conventional
Box-Cox and Yeo-Johnson transformations are well-known. However, the
introduction of scaling parameter \(s\) prevents their direct use. Here,
we first derive the general form of the log-likelihood functions, and
then derive their power-transformation specific definitions.

@@

\begin{equation}
\begin{split}
\mathcal{l}_{\text{BC}}^{\lambda, x_0, s} = & -\frac{n}{2} \log \left[2 \pi \sigma^2 \right] -\frac{1}{2 \sigma^2} \sum_{i=1}^n \left( \phi^{\lambda, x_0, s}(x_i) - \mu \right)^2 \\
& -n \lambda \log s + \left( \lambda - 1 \right) \sum_{i=1}^n \log \left[ x_i - x_0 \right]
\end{split}
\end{equation}

Similarly to Raymaekers and Rousseeuw (2024), sample weights \(w_i\) are
introduced to facilitate robust power transformations. The weighted
log-likelihood of the location- and scale-invariant Box-Cox
transformation is:

\begin{equation}
\begin{split}
\mathcal{l}_{\text{rBC}}^{\lambda, x_0, s} = & -\frac{1}{2} \left(\sum_{i=1}^n w_i \right) \log \left[ 2 \pi \sigma_w^2 \right] -\frac{1}{2 \sigma_w^2} \sum_{i=1}^n w_i \left( \phi^{\lambda, x_0, s}(x_i) - \mu_w \right)^2 \\
& - \lambda \left( \sum_{i=1}^n w_i \right) \log s + \left( \lambda - 1 \right) \sum_{i=1}^n w_i \log \left[ x_i - x_0 \right]
\end{split}
\end{equation}

@@

Let \(\mathbf{X}\) be a feature with ordered feature values, and
\(\mathbf{Y}^\lambda =\phi^{\lambda} \left(\mathbf{X} \right)\) and
\(\mathbf{Y}^{\lambda, x_0, s} =\phi^{\lambda, x_0, s} \left(\mathbf{X} \right)\)
its transformed values using conventional and shift and scale invariant
power transformations, respectively. Since power transformations are
monotonic, \(\mathbf{Y}\) will likewise be ordered.

Below we will focus on criteria based on the empirical density function
and those based on skewness and kurtosis of the transformed featured.
Other potential criteria, such as the Shapiro-Wilk test statistic
(Shapiro and Wilk 1965) are not investigated here. In the case of the
Shapiro-Wilk test statistic this is because of lack of scalability to
features with many (\(> 5000\)) instances, and because adapting the test
statistic to include weights is not straightforward.

@@

Here \(w_i\) are weights, and \(\mu\) and \(\sigma\) are location and
scale parameters. For non-robust power transformations, all \(w_i = 1\).
Note that this criterion is not the same as the Anderson-Darling test
statistic (Anderson and Darling 1952), which involves solving (or
approximating) an integral function, contains an extra scalar
multiplication term, and does not include weights. The Anderson-Darling
criterion seeks to minimise the squared Euclidean distance between the
EDF and the normal CDF, with differences at the upper and lower end of
the normal CDF receiving more weight than those at the the centre of the
CDF.

@@

Here \(w_i\) are weights, and \(\mu\) and \(\sigma\) are location and
scale parameters. For non-robust power transformations, all \(w_i = 1\).
The criterion is similar to the Cramér-von Mises test statistic Mises
(1928), aside from a additive scalar value and the introduction of
weights. This criterion, like the Anderson-Darling criterion, seeks to
minimise the squared Euclidean distance between the EDF and the normal
CDF. Unlike the Anderson-Darling criterion, this criterion weights all
instances equally.

@@

The second class of criteria seeks to reduce skewness and (excess)
kurtosis of the transformed feature \(\mathbf{Y}\). We will first define
the location \(\mu\) and scale \(\sigma\) of the the transformed as
these are required for computing skewness and kurtosis. Here, \(\mu\) is
defined as:

@@

The location, or mean, is weighted using weights \(w_i\). For non-robust
transformations, \(w_i = 1\). Then, \(\sigma^2\) is defined as:

@@

The D'Agostino criterion defined here follows the D'Agostino \(K^2\)
test statistic (D'Agostino and Belanger 1990). This test statistic is
composed of two separate test statistics, one of which is related to
skewness, and the other to kurtosis. Both test statistics are computed
in several steps. Let us first define \(\nu=\sum_{i=1}^n w_i\). Thus for
non-robust power transformations, \(\nu = n\).

For the skewness test statistic we first compute (D'Agostino and
Belanger 1990):

@@

For the kurtosis test statistic we first compute (D'Agostino and
Belanger 1990; Anscombe and Glynn 1983):

@@

The main difference between the above criterion and the Jarque-Bera test
statistic (Jarque and Bera 1980) is that a scalar multiplication is
absent.

@@

\section{Appendix D: Central normality test and empirical central
normality test critical
values}\label{appendix-d-central-normality-test-and-empirical-central-normality-test-critical-values}

Critical values for the central normality test and its empirical variant
are found in Table \ref{tab:central-normality-critical-statistic} and
Table \ref{tab:empirical-central-normality-critical-statistic},
respectively.

\begin{table}
\begin{center}
\caption{Critical values of test statistic $\tau_{\alpha, n, \kappa = 0.80}$ for
\textbf{central normality} at $\kappa = 0.80$, as a function of significance 
level $\alpha$ and number of instances $n$. The values shown must be divided by $100$.}
\label{tab:central-normality-critical-statistic}
\begin{tabular}{l | r r r r r r r r r}

@@

\midrule
    5 & 210.79 & 95.75 & 66.27 & 49.69 & 35.93 & 27.60 & 19.29 & 13.41 & 11.00 \\
   10 &  48.04 & 31.44 & 26.96 & 23.91 & 21.11 & 18.16 & 13.66 & 10.14 &  8.67 \\
   20 &  29.13 & 22.13 & 19.75 & 17.94 & 16.00 & 13.81 & 10.54 &  8.18 &  7.18 \\
   50 &  17.31 & 14.01 & 12.63 & 11.48 & 10.28 &  9.00 &  7.00 &  5.54 &  4.94 \\
  100 &  12.19 & 10.04 &  9.02 &  8.21 &  7.38 &  6.47 &  5.04 &  3.99 &  3.56 \\
  200 &   8.68 &  7.11 &  6.40 &  5.83 &  5.22 &  4.57 &  3.59 &  2.86 &  2.56 \\ 
  500 &   5.61 &  4.54 &  4.08 &  3.70 &  3.33 &  2.91 &  2.28 &  1.82 &  1.63 \\
 1000 &   3.93 &  3.19 &  2.87 &  2.60 &  2.33 &  2.05 &  1.61 &  1.29 &  1.15 \\
 2000 &   2.71 &  2.24 &  2.02 &  1.84 &  1.66 &  1.46 &  1.14 &  0.91 &  0.82 \\
 5000 &   1.73 &  1.42 &  1.28 &  1.17 &  1.05 &  0.92 &  0.72 &  0.58 &  0.52 \\
10000 &   1.22 &  1.00 &  0.90 &  0.82 &  0.74 &  0.65 &  0.51 &  0.41 &  0.37 \\
\bottomrule
\end{tabular}
\end{center}
\end{table}

\begin{table}
\begin{center}
\caption{Critical values of test statistic $\tau_{\alpha, n, \kappa = 0.70}$ for
\textbf{empirical central normality} at $\kappa = 0.70$, as a function of significance 
level $\alpha$ and number of instances $n$. The values shown must be divided by $100$.}
\label{tab:empirical-central-normality-critical-statistic}
\begin{tabular}{l | r r r r r r r r r}

\midrule
    5 & 159.27 & 66.78 & 57.06 & 50.55 & 43.42 & 34.89 & 22.28 & 14.92 & 12.10 \\
   10 &  41.82 & 30.82 & 27.01 & 24.31 & 21.58 & 18.53 & 13.82 & 10.23 &  8.75 \\
   20 &  29.44 & 23.65 & 21.23 & 19.16 & 17.01 & 14.67 & 11.10 &  8.42 &  7.35 \\
   50 &  18.95 & 15.56 & 13.96 & 12.74 & 11.35 &  9.82 &  7.51 &  5.83 &  5.16 \\ 
  100 &  13.49 & 11.00 &  9.91 &  8.98 &  8.03 &  7.04 &  5.43 &  4.25 &  3.76 \\
  200 &   9.65 &  7.99 &  7.22 &  6.64 &  5.96 &  5.21 &  4.05 &  3.17 &  2.81 \\
  500 &   6.37 &  5.32 &  4.82 &  4.44 &  4.03 &  3.57 &  2.79 &  2.18 &  1.92 \\
 1000 &   4.64 &  3.95 &  3.65 &  3.39 &  3.11 &  2.78 &  2.20 &  1.73 &  1.53 \\
 2000 &   3.61 &  3.18 &  2.96 &  2.78 &  2.57 &  2.32 &  1.89 &  1.50 &  1.33 \\
 5000 &   2.78 &  2.48 &  2.36 &  2.24 &  2.11 &  1.95 &  1.66 &  1.39 &  1.25 \\
10000 &   2.39 &  2.18 &  2.08 &  2.00 &  1.91 &  1.81 &  1.59 &  1.39 &  1.28 \\
\bottomrule
\end{tabular}
\end{center}
\end{table}

{\centering \includegraphics{manuscript_appendix_files/figure-latex/empirical-central-normality-critical-statistics-1}

\caption{Critical test statistic $\tau_{\alpha, n, \kappa = 0.80}$ of the (empirical) central normality test as function of $n$ for several values of significance level $\alpha$. The critical test statistics for central normality test are determined using fully normal data, whereas the statistics for the empirical variant are determined using centrally normal data, i.e. with fully normal data where 10\% of elements are replaced by outliers. emp: empirical}\label{fig:empirical-central-normality-critical-statistics}
\end{figure}

@@

\section{Appendix E: Assessing transformations using simulated
data}\label{appendix-e-assessing-transformations-using-simulated-data}

@@

\begin{table}
\begin{center}
\caption{
Comparison of average rank between Box-Cox transformation methods based on either residual error (without outliers) or residual error of the central portion
(with outliers; $\kappa = 0.80$) over 3 datasets with 10000 sequences each. The clean dataset consists of sequences derived through inverse Box-Cox transformation
of data sampled from a standard normal distribution. The dirty dataset contains sequences sampled from asymmetric generalised normal distributions.
The shifted dataset also contains sequences sampled from asymmetric generalised normal distributions, but centred at 100, and scaled by 0.001.
If necessary, each sequence was shifted so that every element had a strictly positive value. 
Several transformation methods include normalisation before transformation, indicated by z-score normalisation (norm.) or robust scaling.
A rank of 1 is the best and a rank of 9 the worst. For each dataset, the best ranking transformation is marked in bold.
}
\label{tab:comparison_methods_simulations_box_cox}
\begin{tabular}{l | l r r r r r r}

@@

none                                  & &         7.65  &         6.96  &         7.28  &         6.58  &         7.45  &         6.78 \\
conventional                          & &         3.10  &         5.98  &         5.42  &         6.16  &         6.49  &         6.10 \\
conventional (z-score norm.)          & &         4.19  &         6.28  &         4.85  &         6.33  &         4.38  &         5.85 \\
conventional (robust scaling)         & &         4.20  &         6.28  &         4.78  &         6.33  &         4.30  &         5.86 \\
Raymaekers-Rousseeuw                  & &         4.46  &         3.55  &         5.39  &         3.69  &         6.29  &         5.84 \\ 
Raymaekers-Rousseeuw (z-score norm.)  & &         6.21  &         3.83  &         4.95  &         3.85  &         4.51  &         3.50 \\ 
Raymaekers-Rousseeuw (robust scaling) & &         6.12  &         3.82  &         4.97  &         3.85  &         4.52  &         3.50 \\
invariant                             & & \textbf{2.99} &         5.55  & \textbf{3.65} &         6.01  & \textbf{3.52} &         5.57 \\
robust invariant                      & &         6.09  & \textbf{2.75} &         3.71  & \textbf{2.20} &         3.55  & \textbf{2.00} \\

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\documentclass[
  a4paper,
]{article}
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\usepackage[]{natbib}
\bibliographystyle{plainnat}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{placeins}
\newtheorem{definition}{Definition}
\DeclareMathOperator*{\argmax}{argmax}
\DeclareMathOperator*{\argmin}{argmin}
\DeclareMathOperator{\sgn}{sgn}
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Appendix},
  pdfauthor={Alex Zwanenburg, Steffen Löck},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Appendix}
\author{Alex Zwanenburg, Steffen Löck}
\date{2025-10-29}

@@

The parameters of these power transformations can be optimised by
maximising the log-likelihood function, under the assumption that the
transformed feature \(\phi^{\lambda, x_0, s} (\mathbf{X})\) follows a
normal distribution. The log-likelihood functions for conventional
Box-Cox and Yeo-Johnson transformations are well-known. However, the
introduction of scaling parameter \(s\) prevents their direct use. Here,
we first derive the general form of the log-likelihood functions, and
then derive their power-transformation specific definitions.

@@

\begin{equation}
\begin{split}
\mathcal{L}_{\text{BC}}^{\lambda, x_0, s} = & -\frac{n}{2} \log \left[2 \pi \sigma^2 \right] -\frac{1}{2 \sigma^2} \sum_{i=1}^n \left( \phi^{\lambda, x_0, s}(x_i) - \mu \right)^2 \\
& -n \lambda \log s + \left( \lambda - 1 \right) \sum_{i=1}^n \log \left[ x_i - x_0 \right]
\end{split}
\end{equation}

Similarly to \citet{Raymaekers2024-zf}, sample weights \(w_i\) are
introduced to facilitate robust power transformations. The weighted
log-likelihood of the location- and scale-invariant Box-Cox
transformation is:

\begin{equation}
\begin{split}
\mathcal{L}_{\text{rBC}}^{\lambda, x_0, s} = & -\frac{1}{2} \left(\sum_{i=1}^n w_i \right) \log \left[ 2 \pi \sigma_w^2 \right] -\frac{1}{2 \sigma_w^2} \sum_{i=1}^n w_i \left( \phi^{\lambda, x_0, s}(x_i) - \mu_w \right)^2 \\
& - \lambda \left( \sum_{i=1}^n w_i \right) \log s + \left( \lambda - 1 \right) \sum_{i=1}^n w_i \log \left[ x_i - x_0 \right]
\end{split}
\end{equation}

@@

Let \(\mathbf{X}\) be a feature with ordered feature values, and
\(\mathbf{Y}^\lambda =\phi^{\lambda} \left(\mathbf{X} \right)\) and
\(\mathbf{Y}^{\lambda, x_0, s} =\phi^{\lambda, x_0, s} \left(\mathbf{X} \right)\)
its transformed values using conventional and location- and
scale-invariant power transformations, respectively. Since power
transformations are monotonic, \(\mathbf{Y}\) will likewise be ordered.

Below we will focus on criteria based on the empirical density function
and those based on skewness and kurtosis of the transformed feature.
Other potential criteria, such as the Shapiro-Wilk test statistic
\citep{Shapiro1965-zd} are not investigated here. In the case of the
Shapiro-Wilk test statistic this is because of lack of scalability to
features with many (\(> 5000\)) instances, and because adapting the test
statistic to include weights is not straightforward.

@@

Here \(w_i\) are weights, and \(\mu\) and \(\sigma\) are location and
scale parameters. For non-robust power transformations, all \(w_i = 1\).
Note that this criterion is not the same as the Anderson-Darling test
statistic \citep{Anderson1952-gz}, which involves solving (or
approximating) an integral function, contains an extra scalar
multiplication term, and does not include weights. The Anderson-Darling
criterion seeks to minimise the squared Euclidean distance between the
EDF and the normal CDF, with differences at the upper and lower end of
the normal CDF receiving more weight than those at the centre of the
CDF.

@@

Here \(w_i\) are weights, and \(\mu\) and \(\sigma\) are location and
scale parameters. For non-robust power transformations, all \(w_i = 1\).
The criterion is similar to the Cramér-von Mises test statistic
\citep[\citet{Von_Mises1928-ef}]{Cramer1928-rc}, aside from an additive
scalar value and the introduction of weights. This criterion, like the
Anderson-Darling criterion, seeks to minimise the squared Euclidean
distance between the EDF and the normal CDF. Unlike the Anderson-Darling
criterion, this criterion weights all instances equally.

@@

The second class of criteria seeks to reduce skewness and (excess)
kurtosis of the transformed feature \(\mathbf{Y}\). We will first define
the location \(\mu\) and scale \(\sigma\) of the transformed feature as
these are required for computing skewness and kurtosis. Here, \(\mu\) is
defined as:

@@

The location, or mean, is weighted by \(w_i\), which equals \(1\) for
non-robust transformations.

Then, \(\sigma^2\) is defined as:

@@

The D'Agostino criterion defined here follows the D'Agostino \(K^2\)
test statistic \citep{DAgostino1990-kp}. This test statistic is composed
of two separate test statistics, one of which is related to skewness,
and the other to kurtosis. Both test statistics are computed in several
steps. Let us first define \(\nu=\sum_{i=1}^n w_i\). Thus for non-robust
power transformations, \(\nu = n\).

For the skewness test statistic we first compute
\citep{DAgostino1990-kp}:

@@

For the kurtosis test statistic we first compute
\citep{DAgostino1990-kp, Anscombe1983-nz}:

@@

The main difference between the above criterion and the Jarque-Bera test
statistic \citep{Jarque1980-hw} is that a scalar multiplication is
absent.

@@

\section{Appendix C: Central normality test and empirical central
normality test critical
values}\label{appendix-c-central-normality-test-and-empirical-central-normality-test-critical-values}

Critical values for the empirical central normality test at
\(\kappa = 0.80\) are found in Table
\ref{tab:empirical-central-normality-critical-statistic}.

\begin{table}
\begin{center}
\caption{Critical values of test statistic $\tau_{\alpha, n, \kappa = 0.80}$ for
central normality at $\kappa = 0.80$, as a function of significance 
level $\alpha$ and number of instances $n$. The values in this table should be divided by $100$ before use.}
\label{tab:empirical-central-normality-critical-statistic}
\begin{tabular}{l | r r r r r r r r r}

@@

\midrule
    5 & 200.88 & 90.25 & 64.46 & 49.04 & 36.07 & 27.70 & 19.28 & 13.41 & 10.90 \\
   10 &  45.90 & 31.78 & 27.12 & 24.14 & 21.23 & 18.30 & 13.68 & 10.20 &  8.71 \\
   20 &  28.88 & 22.53 & 19.96 & 18.01 & 15.99 & 13.93 & 10.67 &  8.26 &  7.27 \\
   50 &  16.86 & 14.10 & 12.66 & 11.55 & 10.28 &  8.95 &  7.00 &  5.53 &  4.91 \\
  100 &  12.41 & 10.00 &  9.02 &  8.21 &  7.36 &  6.45 &  5.02 &  4.00 &  3.58 \\
  200 &   8.65 &  7.05 &  6.37 &  5.80 &  5.22 &  4.57 &  3.57 &  2.84 &  2.54 \\
  500 &   5.54 &  4.47 &  4.03 &  3.69 &  3.30 &  2.90 &  2.27 &  1.81 &  1.63 \\
 1000 &   3.88 &  3.17 &  2.86 &  2.60 &  2.34 &  2.05 &  1.61 &  1.28 &  1.15 \\
 2000 &   2.79 &  2.24 &  2.03 &  1.84 &  1.66 &  1.45 &  1.14 &  0.91 &  0.82 \\ 
 5000 &   1.75 &  1.41 &  1.27 &  1.17 &  1.05 &  0.92 &  0.72 &  0.58 &  0.52 \\
10000 &   1.21 &  1.00 &  0.90 &  0.82 &  0.74 &  0.65 &  0.51 &  0.41 &  0.36 \\
\bottomrule
\end{tabular}
\end{center}
\end{table}

@@

\section{Appendix D: Assessing transformations using simulated
data}\label{appendix-d-assessing-transformations-using-simulated-data}

@@

\begin{table}
\begin{center}
\caption{
Comparison of average rank between Box-Cox transformation methods based on either residual error (without outliers) or residual error of the central portion
(with outliers; $\kappa = 0.80$) over 3 datasets with 100000 sequences each. The clean dataset consists of sequences derived through inverse Box-Cox transformation
of data sampled from a standard normal distribution. The dirty dataset contains sequences sampled from asymmetric generalised normal distributions.
The shifted dataset also contains sequences sampled from asymmetric generalised normal distributions, but centred at 100, and scaled by 0.001.
If necessary, each sequence was shifted so that every element had a strictly positive value. 
Several transformation methods include normalisation before transformation, indicated by z-score normalisation (norm.) or robust scaling.
A rank of 1 is the best and a rank of 9 the worst. For each dataset, the best ranking transformation is marked in bold.
}
\label{tab:comparison_methods_simulations_box_cox}
\begin{tabular}{l | l r r r r r r}

@@

none                                  & &         7.63  &         6.95  &         7.29  &         6.59  &         7.47  &         6.78 \\
conventional                          & &         3.11  &         5.98  &         5.41  &         6.18  &         6.51  &         6.10 \\
conventional (z-score norm.)          & &         4.15  &         6.27  &         4.83  &         6.32  &         4.34  &         5.84 \\
conventional (robust scaling)         & &         4.20  &         6.29  &         4.84  &         6.31  &         4.35  &         5.84 \\
Raymaekers-Rousseeuw                  & &         4.50  &         3.56  &         5.39  &         3.69  &         6.30  &         5.80 \\ 
Raymaekers-Rousseeuw (z-score norm.)  & &         6.23  &         3.80  &         4.93  &         3.85  &         4.48  &         3.50 \\ 
Raymaekers-Rousseeuw (robust scaling) & &         6.12  &         3.80  &         4.99  &         3.87  &         4.53  &         3.51 \\
invariant                             & & \textbf{2.96} &         5.59  & \textbf{3.62} &         6.01  & \textbf{3.51} &         5.58 \\
robust invariant                      & &         6.10  & \textbf{2.76} &         3.68  & \textbf{2.19} & \textbf{3.51} & \textbf{2.00} \\

\subsection{Ranking Yeo-Johnson transformations: residual
errors}\label{ranking-yeo-johnson-transformations-residual-errors}

In the main manuscript, transformation methods are compared by ranking
within each experiment, defined by the combination of sequence and
absence/presence of outliers. This allows for direct comparison of
methods between different experiments.

Table \ref{tab:comparison_methods_simulations_yeo_johnson_avg_residual}
shows the average normalised residual error for these transformation
methods. This metric is computed from the residual error and the
residual error of the central portion (\(\kappa = 0.70\)) for sequences
without and with outliers, respectively. In each experiment, the
respective residual error was normalised by dividing with the sequence
length, and can be interpreted as the average residual error of a single
sample. This reduces the effect of sequence length on the residual
error.

Table
\ref{tab:comparison_methods_simulations_yeo_johnson_avg_residual_diff}
shows the average normalised difference in residual error between the
best method for each experiment and the individual methods. As above,
normalisation was done by division by the length of the sequence
corresponding to each experiment.

Both tables show that the differences between best-performing methods on
each dataset are relatively small.

\begin{table}
\begin{center}
\caption{
Comparison of average normalised residual error between Yeo-Johnson transformation methods based on either the residual error (without outliers) or the residual error of the central portion
(with outliers; $\kappa = 0.80$) over 3 datasets with 100000 sequences each. Residual errors were normalised by sequence length to reduce the effect of sequence length on the residual error.
The clean dataset consists of sequences derived through inverse Yeo-Johnson transformation
of data sampled from a standard normal distribution. The dirty dataset contains sequences sampled from asymmetric generalised normal distributions, centred at 0.
The shifted dataset also contains sequences sampled from asymmetric generalised normal distributions, but centred at 100, and scaled by 0.001.
Several transformation methods include normalisation before transformation, indicated by z-score normalisation (norm.) or robust scaling.
A lower normalised residual error is better. Discrepancies with the rank-based method may occur, as even with normalisation residual errors 
are not fully comparable between experiments, whereas the ranks achieved within each experiment are.
}
\label{tab:comparison_methods_simulations_yeo_johnson_avg_residual}
\begin{tabular}{l | l r r r r r r}

none                                  & & 0.202 & 0.112 & 0.128 & 0.080 & 0.129 & 0.080 \\
conventional                          & & 0.058 & 0.094 & 0.085 & 0.078 & 0.129 & 0.080 \\
conventional (z-score norm.)          & & 0.061 & 0.094 & 0.091 & 0.073 & 0.091 & 0.073 \\
conventional (robust scaling)         & & 0.058 & 0.094 & 0.090 & 0.073 & 0.090 & 0.073 \\
Raymaekers-Rousseeuw                  & & 0.071 & 0.055 & 0.092 & 0.279 & 0.129 & 0.080 \\
Raymaekers-Rousseeuw (z-score norm.)  & & 0.087 & 0.057 & 0.097 & 0.060 & 0.097 & 0.060 \\
Raymaekers-Rousseeuw (robust scaling) & & 0.071 & 0.056 & 0.095 & 0.059 & 0.095 & 0.059 \\
invariant                             & & 0.057 & 0.099 & 0.083 & 0.075 & 0.083 & 0.075 \\
robust invariant                      & & 0.069 & 0.066 & 0.084 & 0.058 & 0.084 & 0.058 \\

\begin{table}
\begin{center}
\caption{
Comparison of average difference in residual error between Yeo-Johnson
transformation methods and the best transformation method within each
experiment, normalised by the corresponding sequence length. This is based on
either the residual error (without outliers) or the residual error of the
central portion (with outliers; $\kappa = 0.80$) over 3 datasets with 100000
sequences each. Residual errors were normalised by sequence length to reduce the
effect of sequence length on the residual error. The clean dataset consists of
sequences derived through inverse Yeo-Johnson transformation of data sampled
from a standard normal distribution. The dirty dataset contains sequences
sampled from asymmetric generalised normal distributions, centred at 0. The
shifted dataset also contains sequences sampled from asymmetric generalised
normal distributions, but centred at 100, and scaled by 0.001. Several
transformation methods include normalisation before transformation, indicated by
z-score normalisation (norm.) or robust scaling. A lower difference is better,
with a minimum value of 0.0. Discrepancies with the rank-based method may occur,
as normalised differences in residual error are not fully comparable between
experiments, whereas the ranks achieved within each experiment are.
}
\label{tab:comparison_methods_simulations_yeo_johnson_avg_residual_diff}
\begin{tabular}{l | l r r r r r r}

none                                  & & 0.150 & 0.068 & 0.050 & 0.033 & 0.049 & 0.033 \\
conventional                          & & 0.005 & 0.049 & 0.006 & 0.032 & 0.049 & 0.033 \\
conventional (z-score norm.)          & & 0.008 & 0.049 & 0.013 & 0.026 & 0.011 & 0.025 \\
conventional (robust scaling)         & & 0.005 & 0.050 & 0.012 & 0.026 & 0.010 & 0.026 \\
Raymaekers-Rousseeuw                  & & 0.018 & 0.011 & 0.014 & 0.233 & 0.049 & 0.033 \\
Raymaekers-Rousseeuw (z-score norm.)  & & 0.034 & 0.012 & 0.018 & 0.014 & 0.017 & 0.013 \\
Raymaekers-Rousseeuw (robust scaling) & & 0.018 & 0.011 & 0.017 & 0.012 & 0.015 & 0.011 \\
invariant                             & & 0.004 & 0.054 & 0.005 & 0.029 & 0.003 & 0.028 \\
robust invariant                      & & 0.016 & 0.021 & 0.006 & 0.011 & 0.004 & 0.010 \\

\subsection{Ranking Yeo-Johnson transformations: effect of sequence
length}\label{ranking-yeo-johnson-transformations-effect-of-sequence-length}
